# Clustering {#ch-clustering}

We now turn to clustering, which is the branch of UL focused on partitioning our data into subgroups of similar observations.  Hereafter, we use the terms clusters and subgroups interchangeably.  Recall from Section \@ref(sec-clustering-hypothesis) of Chapter \@ref(ch-goals) that clustering rests on the assumption that our data organize into well defined subsets, which raises two linked questions: how many clusters (if any) exist, and what do we mean by similar observations?  As the visual examples there showed, the answer to the second question -- our notion of similarity -- directly shapes the answer to the first, and this will not be so easy to resolve visually once we move beyond two dimensions.  We now discuss data driven ways to estimate the appropriate number of clusters for a specific algorithm.

There is one main commonality (and associated set of notation) that will persist throughout this section.  The ultimate goal of clustering is to separate data $\vec{x}_1,\dots,\vec{x}_N$ into $k$ groups. The choice of $k$ is critical in clustering.  Some algorithms proceed by iteratively merging observations until only $k$ clusters remain (hierarchical methods). Others (center- and model-based and spectral methods) treat clustering as a partitioning problem, where each element of the partition corresponds to a pre-specified number of clusters. Algorithm dependent methods for choosing $k$ can be used for deciding optimal number of clusters without a priori knowledge.


<!-- Center based clustering -->
```{r child = 'topics/Center-Based-Clustering.Rmd'}
```

## Kernel $k$-means {#sec-kernel-kmeans}

The kernel trick is not limited to PCA.  Recall from \@ref(ch-nonlinear) that the kernel trick lets us replace inner products $\vec{x}_i^T\vec{x}_j$ with a kernel evaluation $k(\vec{x}_i,\vec{x}_j)$, giving many linear methods access to nonlinear structure without ever explicitly computing the feature map $\varphi$.  The $k$-means loss function introduced above is a natural candidate, since expanding the squared norm shows that it only involves inner products of the data,
$$\|\vec{x}_i - \vec{\mu}_\ell\|^2 = \vec{x}_i^T\vec{x}_i - \frac{2}{|C_\ell|}\sum_{\vec{x}_j \in C_\ell} \vec{x}_i^T\vec{x}_j + \frac{1}{|C_\ell|^2}\sum_{\vec{x}_j,\vec{x}_{j'} \in C_\ell} \vec{x}_j^T\vec{x}_{j'}, \qquad \vec{\mu}_\ell = \frac{1}{|C_\ell|}\sum_{\vec{x}_j \in C_\ell} \vec{x}_j.$$
Replacing each inner product $\vec{x}_i^T\vec{x}_j$ with a kernel evaluation $k(\vec{x}_i,\vec{x}_j)$ gives the squared distance from $\varphi(\vec{x}_i)$ to the (never explicitly computed) feature-space cluster mean $\bar\varphi_\ell = \frac{1}{|C_\ell|}\sum_{\vec{x}_j \in C_\ell} \varphi(\vec{x}_j)$,
$$\|\varphi(\vec{x}_i) - \bar\varphi_\ell\|^2_{\mathcal{H}} = k(\vec{x}_i,\vec{x}_i) - \frac{2}{|C_\ell|}\sum_{\vec{x}_j \in C_\ell} k(\vec{x}_i,\vec{x}_j) + \frac{1}{|C_\ell|^2}\sum_{\vec{x}_j,\vec{x}_{j'} \in C_\ell} k(\vec{x}_j,\vec{x}_{j'}).$$

Kernel $k$-means then proceeds just like Lloyd's Algorithm above, but replaces the assignment step's (squared) Euclidean distance with this kernelized distance.

1. **Initialize**: Choose an initial partition $C_1,\dots,C_k$ (e.g. by randomly assigning points to $k$ groups).
2. **Cluster Assignment**: Reassign each $\vec{x}_i$ to the cluster $\ell$ minimizing $\|\varphi(\vec{x}_i)-\bar\varphi_\ell\|^2_{\mathcal{H}}$ given the current partition.
3. **Repeat** step 2, recomputing the expression above from the updated partition, until cluster assignments stop changing.

Notice there is no analog of the "center update" step from ordinary $k$-means.  Since $\bar\varphi_\ell$ typically cannot be computed or even represented explicitly, the cluster membership itself is the only state carried between iterations.

As with kernel PCA (\@ref(ch-nonlinear)), the choice and tuning of the kernel is critical.  A poorly tuned kernel gives a result no better (and no worse) than ordinary $k$-means, while a well chosen kernel can recover cluster structure that is not spherical in the original coordinates.

::: {.example #kkmeans-rings name="Kernel $k$-means applied to Nested Rings"}

Consider $N=400$ points sampled uniformly around two concentric circles with additive noise, a shape ordinary $k$-means cannot separate correctly since neither ring is convex (let alone spherical).

```{r, echo = FALSE, message = FALSE, fig.cap = "Nested rings which are not separable by ordinary k-means"}
library("kernlab")
set.seed(185)
N <- 400
theta <- runif(N, 0, 2*pi)
r <- rep(c(1,3), each = N/2) + rnorm(N, sd = 0.1)
rings <- cbind(r*cos(theta), r*sin(theta))
plot(rings, xlab = "", ylab = "", pch = 16, cex = 0.5, asp = 1)
```

Below, we compare the partition found by ordinary $k$-means to the partition found by kernel $k$-means using an rbf kernel.

```{r, echo = FALSE, fig.cap = "Ordinary k-means (left) vs. kernel k-means with an rbf kernel (right)"}
out.km <- kmeans(rings, centers = 2)
out.kkm <- kkmeans(rings, centers = 2, kernel = "rbfdot", kpar = list(sigma = 4))
par(mfrow = c(1,2))
plot(rings, col = out.km$cluster, main = "k-means", xlab = "", ylab = "", asp = 1)
plot(rings, col = out.kkm, main = "Kernel k-means (rbf)", xlab = "", ylab = "", asp = 1)
```

Ordinary $k$-means splits the data by a hyperplane, cutting each ring in half.  Kernel $k$-means, by first (implicitly) mapping the data to a feature space where the rings become linearly separable, recovers the two rings exactly.

:::

Kernel $k$-means also has an important connection to a clustering method discussed later in this chapter: with an appropriately chosen (graph-based) kernel, minimizing the kernel $k$-means loss is equivalent to minimizing a normalized graph cut, which is exactly the objective used by spectral clustering (\@ref(sec-spec-clustering)) [@dhillon_kkmeans].  In other words, kernel $k$-means and spectral clustering are two algorithmic routes to the same underlying idea.

<!-- Hierarchical -->
```{r child = 'topics/Hierarchical-Clustering.Rmd'}
```

<!-- Model based -->
```{r child = 'topics/Model-Based-Clustering.Rmd'}
```

<!-- Spectral Clustering -->
```{r child = 'topics/Spectral-Clustering.Rmd'}
```

<!-- DBSCAN -->
```{r child = 'topics/DBSCAN.Rmd'}
```

<!-- Clustering comparisons and metrics -->
```{r child = 'topics/Methods-for-Comparing-Clustering.Rmd'}
```

## Summary

If Chapter \@ref(ch-nonlinear) was organized around two strategies for handling curved structure, this chapter is organized around competing geometric definitions of the word "cluster" itself, and choosing a clustering method is really choosing which definition matches your data.

Center-based methods ($k$-means, $k$-medoids, $k$-center) define a cluster as a set of points close to a single representative center, which implicitly commits to roughly ball-shaped clusters (\@ref(sec-center-based-clustering)). Kernel $k$-means keeps this same center-based picture but relocates it to an implicit feature space, letting it recover clusters -- like the nested rings -- that are not ball-shaped in the original coordinates but become so after an implicit nonlinear mapping (\@ref(sec-kernel-kmeans)). Hierarchical clustering drops the notion of a center entirely and instead builds structure bottom-up from pairwise distances, with the choice of linkage function encoding a different geometric rule for how two whole clusters, not just two points, should be compared (\@ref(sec-hierarchical-clustering)). Model-based clustering (GMMs) replaces hard geometric boundaries with soft, probabilistic ones, representing each cluster as an ellipsoidal region of high density and allowing points to belong fractionally to more than one cluster (\@ref(sec-model-based-clustering)). Spectral clustering steps back from distances in the original space altogether, instead defining clusters as well-connected regions of a similarity graph -- a picture that, as we saw, is secretly equivalent to kernel $k$-means with a graph-based kernel (\@ref(sec-spec-clustering)). And DBSCAN commits to yet another picture, defining a cluster as a connected region of high density with no assumption on shape at all, and -- unlike every other method in this chapter -- allowing points to belong to no cluster whatsoever (\@ref(sec-dbscan)).

None of these definitions is more "correct" than the others; each is simply a different geometric lens, well suited to some datasets and poorly suited to others, which is exactly why the nested rings example recurs throughout this chapter as a stress test. Since we generally don't know in advance which geometric picture fits our data, the cluster validation tools at the end of the chapter (\@ref(sec-cluster-validation)) become essential: they let us check, after the fact, whether the geometric assumptions a method made were actually the right ones for the data we gave it.

## Exercises

1. Recall that the intracluster variation is defined as $V(C_\ell) = \frac{1}{2|C_\ell|}\sum_{\vec{x}_i,\vec{x}_j \in C_\ell}\|\vec{x}_i-\vec{x}_j\|^2$. Show algebraically that $$V(C_\ell) = \sum_{\vec{x}_i \in C_\ell}\|\vec{x}_i-\vec{\mu}_\ell\|^2$$ where $\vec{\mu}_\ell$ is the sample mean of the points in $C_\ell$. Hint: expand $\|\vec{x}_i-\vec{x}_j\|^2 = \|(\vec{x}_i-\vec{\mu}_\ell)-(\vec{x}_j-\vec{\mu}_\ell)\|^2$ and use the fact that $\sum_{\vec{x}_i\in C_\ell}(\vec{x}_i-\vec{\mu}_\ell) = \vec{0}$.

1. Let $\vec{x}_1,\dots,\vec{x}_4 = 0,1,9,10 \in\mathbb{R}$ and consider two candidate partitions into $k=2$ clusters: Partition A, $C_1=\{0,1\}, C_2=\{9,10\}$; and Partition B, $C_1=\{0,9\}, C_2=\{1,10\}$.

    a. Compute $T_{kmeans}$ for Partition A and for Partition B.

    b. Which partition does the $k$-means loss prefer? Does this match your intuition?

1. Let $\vec{x}_1,\dots,\vec{x}_5 = -2,-1,0,1,10 \in \mathbb{R}$, and consider a single cluster ($k=1$) containing all five points.

    a. Find the center $c$ minimizing the $k$-means loss $\sum_i (x_i-c)^2$, and compute the resulting loss.

    b. Find the center $c\in\{x_1,\dots,x_5\}$ minimizing the $k$-medoids loss $\sum_i |x_i-c|$, and compute the resulting loss.

    c. The outlier at $x=10$ pulls the $k$-means center far from the tight cluster of the other four points, while the $k$-medoids center stays inside it. Explain why, using the difference between squared and non-squared distance.

1. Consider $\vec{x}_1,\dots,\vec{x}_6 = 1,2,3,8,9,10\in\mathbb{R}$ and $k=2$. Run Lloyd's algorithm by hand starting from initial centers $c_1=1,\, c_2=4$.

    a. Find the cluster assignments and updated centers after one iteration.

    b. Find the cluster assignments and updated centers after a second iteration.

    c. Has the algorithm converged? How does the final partition compare to what you'd guess visually?

1. Consider four points in $\mathbb{R}^2$: $A=(0,0),\, B=(1,0)$ in cluster 1, and $C=(10,0),\, D=(11,0)$ in cluster 2, using Euclidean distance.

    a. Compute the silhouette $s_A$ for point $A$.

    b. Compute the silhouette $s_B$ for point $B$.

    c. Without further computation, state $s_C$ and $s_D$ and explain why symmetry gives you these values for free.

1. Suppose we fit GMMs with $K=1,2,3$ clusters to a dataset with $N=100$ observations, obtaining maximized log-likelihoods $\log L_1^* = -150,\, \log L_2^* = -120,\, \log L_3^*=-115$, and suppose the corresponding number of free parameters are $\mathcal{P}_1=5,\,\mathcal{P}_2=11,\,\mathcal{P}_3=17$.

    a. Compute $BIC_1, BIC_2,$ and $BIC_3$ using $BIC_k = 2\log L_k^* - \mathcal{P}_k \log N$.

    b. Which $K$ does BIC select? Explain why this need not be the $K$ with the largest $\log L_k^*$.

1. For data in $\mathbb{R}^3$ ($d=3$) with $K=4$ clusters, count the number of free parameters (means, covariance parameters, and cluster probabilities combined) under

    a. the EII model (equal volume, spherical shape, no orientation to speak of), i.e. ${\bf \Sigma}_j = \lambda {\bf I}$ for all $j$.

    b. the VVV model (fully unconstrained covariances).

1. Given cluster densities $f_1 = \mathcal{N}(0,1)$ and $f_2 = \mathcal{N}(5,1)$ (both univariate) with equal prior probabilities $p_1=p_2=0.5$, compute the posterior probability $p(Z=1\mid x=2)$ that a point at $x=2$ belongs to cluster 1. Recall from Example \@ref(exm:gmm) that this posterior is what allows GMMs to produce "fuzzy" cluster assignments.

1. Recall the kernel $k$-means distance $\|\varphi(\vec{x}_i)-\bar\varphi_\ell\|_{\mathcal{H}}^2 = k(\vec{x}_i,\vec{x}_i) - \frac{2}{|C_\ell|}\sum_{\vec{x}_j\in C_\ell}k(\vec{x}_i,\vec{x}_j) + \frac{1}{|C_\ell|^2}\sum_{\vec{x}_j,\vec{x}_{j'}\in C_\ell}k(\vec{x}_j,\vec{x}_{j'})$. Show that using the linear kernel $k(\vec{x},\vec{y})=\vec{x}^T\vec{y}$ recovers exactly the ordinary $k$-means distance $\|\vec{x}_i-\vec{\mu}_\ell\|^2$, so that kernel $k$-means with a linear kernel is just ordinary $k$-means.

1. Consider four points $A,B,C,D$ with pairwise distances $d(A,B)=2,\, d(A,C)=8,\, d(A,D)=9,\, d(B,C)=7,\, d(B,D)=8,\, d(C,D)=3$.

    a. Build the dendrogram using **single** linkage. At what heights do the merges occur?

    b. Build the dendrogram using **complete** linkage. At what heights do the merges occur?

    c. The two dendrograms agree on which points merge first, but disagree on the height of the final merge. Explain this difference in terms of the definitions of single and complete linkage.

1. Using the single-linkage dendrogram from the previous exercise:

    a. Write down the cophenetic distance (Definition \@ref(def:cc)) between each of the 6 pairs of points.

    b. Using R, compute the cophenetic correlation coefficient between the original distances and the cophenetic distances.

1. Let $C=\{\{1,2,3\},\{4,5\}\}$ be a clustering of 5 points and let $C'=\{\{1,2\},\{3,4,5\}\}$ be a reference partition (e.g. true labels).

    a. Build the $2\times 2$ contingency table with entries $n_{ij}$ as in Definition \@ref(def:ari).

    b. By directly examining all $\binom{5}{2}=10$ pairs, count how many are concordant (both partitions agree the pair is together, or agree it is apart) and compute the Rand Index.

    c. Using your contingency table from (a) and the formula $RI(C,C') = \left[2\sum_{ij}\binom{n_{ij}}{2} + \binom{N}{2} - \sum_i\binom{n_{i\cdot}}{2}-\sum_j\binom{n_{\cdot j}}{2}\right]\big/\binom{N}{2}$, recompute the Rand Index and confirm it matches your answer to (b).

    d. Compute the Adjusted Rand Index using Definition \@ref(def:ari).

    e. Using a calculator or R, compute the Normalized Mutual Information using Definition \@ref(def:nmi).

1. Two independent, uniformly random partitions of $N=1000$ points, each into 50 clusters of 20 points, tend to have a Rand Index well above zero even though they share no real structure. Explain why, without computing anything -- think about what fraction of pairs of points end up in *different* clusters under each partition, and how that fraction alone drives most of the agreement.

1. Let $\vec{x}_1,\dots,\vec{x}_4 = 0,1,5,6\in\mathbb{R}$, and build the (unweighted, binary) neighborhood graph for spectral clustering using $\epsilon = 2$, so that $\vec{x}_i,\vec{x}_j$ are connected exactly when $|\vec{x}_i-\vec{x}_j|\le \epsilon$.

    a. Write down the adjacency matrix ${\bf W}$ and graph Laplacian ${\bf L}={\bf D}-{\bf W}$.

    b. Find the eigenvalues of ${\bf L}$. How many are exactly zero, and how does this relate to the number of connected components of the graph?

    c. Give a pair of orthogonal eigenvectors spanning the null space of ${\bf L}$, and explain how using these as a 2-dimensional embedding followed by $k$-means with $k=2$ recovers the two connected components exactly.

1. In the Kernel $k$-means section of this chapter, ordinary $k$-means fails on the nested rings example while kernel $k$-means with an rbf kernel succeeds. Using the equivalence between kernel $k$-means (with a graph-based kernel) and spectral clustering, explain conceptually why spectral clustering would also be expected to succeed on the same nested rings data.

1. The `iris` dataset contains known species labels for $N=150$ flowers, giving us external ground truth to validate against.

    a. Standardize the four numeric iris features, then cluster the standardized data into $k=3$ groups using $k$-means, hierarchical clustering (try both average and Ward linkage), and a Gaussian mixture model.

    b. For each of the four clusterings, compute the silhouette score, the Adjusted Rand Index, and the Normalized Mutual Information against the true species labels.

    c. Does the method with the best silhouette score also have the best ARI/NMI? If not, what does that tell you about relying on internal validation alone?

    d. It is well known that two of the three iris species (versicolor and virginica) overlap substantially in feature space, while the third (setosa) is well separated from the other two. Does this show up in your results? Which pairs of species are most often confused by your clusterings?


