# Clustering {#ch-clustering}

We now turn to clustering, which is the branch of UL focused on partitioning our data into subgroups of similar observations.  Hereafter, we use the terms clusters and subgroups interchangeably.  There are naturally a number of questions which arise including:

i) how many clusters (if any) exist?
ii) what do we mean by similar observations?

As we will see through a series of examples, these two points are linked. Given a notion of similarity (which are implicitly defined for different clustering algorithms and tuneable for others), a certain number of clusters and clustering of data may be optimal. Given a different notion of similarity, an entirely different organization of the data into a different number of clusters may arise. 

For now, let us focus on the latter question regarding similarity.  Consider the following three examples of data sets in $\mathbb{R}^2$.  Visually, how would you cluster the observations? In particular, how many subgroups would you say exist and how would your partition the data into these subgroups?

::: {.example #clusterwhat name="Different notions of similar subgroups"}

```{r ex-cluster, fig.align='center', fig.width=10, echo = FALSE}
set.seed(185)
library("MASS")
N <- 200
rho <- 3/4
data1 <- matrix(rnorm(N*2, mean = -2), nrow = N) + matrix((runif(N) < 0.5),nrow = N) %*% matrix(c(4,4),nrow = 1)
data2 <- rbind(mvrnorm(N/2, mu = c(-2,-2), Sigma = matrix(c(1,rho,rho,1),nrow = 2)),
               mvrnorm(N/2, mu = c(2,2), Sigma = matrix(c(1,-rho,-rho,1),nrow = 2)))
data3 <- matrix(rnorm(N*2), nrow = N)
data3 <- diag(1 + (runif(N) < 0.5) + (runif(N) < 0.5)) %*% diag(1/sqrt(diag(data3 %*% t(data3)))) %*% data3
par(mfrow = c(1,3))
plot(data1, xlab = '', ylab = '', main = "Case 1")
plot(data2, xlab = '', ylab = '', main = "Case 2")
plot(data3, xlab = '', ylab = '', main = "Case 3")
```
:::

The questions posed above can be reasonably answered based on the figures but will not be so easy in high dimensions.  We'll explore these questions and discuss data driven ways to estimate the appropriate number of cluster **for a specific algorithm** in the following sections.

There is one main commonality (and associated set of notation) that will persist throughout this section.  The ultimate goal of clustering is to separate data $\vec{x}_1,\dots,\vec{x}_N$ into $k$ groups. The choice of $k$ is critical in clustering.  Some algorithms proceed by iteratively merging observations until only $k$ clusters remain (hierarchical methods). Others (center- and model-based and spectral methods) treat clustering as a partitioning problem, where each element of the partition corresponds to a pre-specified number of clusters. Algorithm dependent methods for choosing $k$ can be used for deciding optimal number of clusters without a priori knowledge.


<!-- Center based clustering -->
```{r child = 'topics/Center-Based-Clustering.Rmd'}
```

## Kernel $k$-means

The kernel trick is not limited to PCA.  Recall from \@ref(ch-nonlinear) that the kernel trick lets us replace inner products $\vec{x}_i^T\vec{x}_j$ with a kernel evaluation $k(\vec{x}_i,\vec{x}_j)$, giving many linear methods access to nonlinear structure without ever explicitly computing the feature map $\varphi$.  The $k$-means loss function introduced above is a natural candidate, since expanding the squared norm shows that it only involves inner products of the data,
$$\|\vec{x}_i - \vec{\mu}_\ell\|^2 = \vec{x}_i^T\vec{x}_i - \frac{2}{|C_\ell|}\sum_{\vec{x}_j \in C_\ell} \vec{x}_i^T\vec{x}_j + \frac{1}{|C_\ell|^2}\sum_{\vec{x}_j,\vec{x}_{j'} \in C_\ell} \vec{x}_j^T\vec{x}_{j'}, \qquad \vec{\mu}_\ell = \frac{1}{|C_\ell|}\sum_{\vec{x}_j \in C_\ell} \vec{x}_j.$$
Replacing each inner product $\vec{x}_i^T\vec{x}_j$ with a kernel evaluation $k(\vec{x}_i,\vec{x}_j)$ gives the squared distance from $\varphi(\vec{x}_i)$ to the (never explicitly computed) feature-space cluster mean $\bar\varphi_\ell = \frac{1}{|C_\ell|}\sum_{\vec{x}_j \in C_\ell} \varphi(\vec{x}_j)$,
$$\|\varphi(\vec{x}_i) - \bar\varphi_\ell\|^2_{\mathcal{H}} = k(\vec{x}_i,\vec{x}_i) - \frac{2}{|C_\ell|}\sum_{\vec{x}_j \in C_\ell} k(\vec{x}_i,\vec{x}_j) + \frac{1}{|C_\ell|^2}\sum_{\vec{x}_j,\vec{x}_{j'} \in C_\ell} k(\vec{x}_j,\vec{x}_{j'}).$$

Kernel $k$-means then proceeds just like Lloyd's Algorithm above, but replaces the assignment step's (squared) Euclidean distance with this kernelized distance.

1. **Initialize**: Choose an initial partition $C_1,\dots,C_k$ (e.g. by randomly assigning points to $k$ groups).
2. **Cluster Assignment**: Reassign each $\vec{x}_i$ to the cluster $\ell$ minimizing $\|\varphi(\vec{x}_i)-\bar\varphi_\ell\|^2_{\mathcal{H}}$ given the current partition.
3. **Repeat** step 2, recomputing the expression above from the updated partition, until cluster assignments stop changing.

Notice there is no analog of the "center update" step from ordinary $k$-means.  Since $\bar\varphi_\ell$ typically cannot be computed or even represented explicitly, the cluster membership itself is the only state carried between iterations.

As with kernel PCA (\@ref(ch-nonlinear)), the choice and tuning of the kernel is critical.  A poorly tuned kernel gives a result no better (and no worse) than ordinary $k$-means, while a well chosen kernel can recover cluster structure that is not spherical in the original coordinates.

::: {.example #kkmeans-rings name="Kernel $k$-means applied to Nested Rings"}

Consider $N=400$ points sampled uniformly around two concentric circles with additive noise, a shape ordinary $k$-means cannot separate correctly since neither ring is convex (let alone spherical).

```{r, echo = FALSE, fig.cap = "Nested rings which are not separable by ordinary k-means"}
library("kernlab")
set.seed(185)
N <- 400
theta <- runif(N, 0, 2*pi)
r <- rep(c(1,3), each = N/2) + rnorm(N, sd = 0.1)
rings <- cbind(r*cos(theta), r*sin(theta))
plot(rings, xlab = "", ylab = "", pch = 16, cex = 0.5, asp = 1)
```

Below, we compare the partition found by ordinary $k$-means to the partition found by kernel $k$-means using an rbf kernel.

```{r, echo = FALSE, fig.cap = "Ordinary k-means (left) vs. kernel k-means with an rbf kernel (right)"}
out.km <- kmeans(rings, centers = 2)
out.kkm <- kkmeans(rings, centers = 2, kernel = "rbfdot", kpar = list(sigma = 4))
par(mfrow = c(1,2))
plot(rings, col = out.km$cluster, main = "k-means", xlab = "", ylab = "", asp = 1)
plot(rings, col = out.kkm, main = "Kernel k-means (rbf)", xlab = "", ylab = "", asp = 1)
```

Ordinary $k$-means splits the data by a hyperplane, cutting each ring in half.  Kernel $k$-means, by first (implicitly) mapping the data to a feature space where the rings become linearly separable, recovers the two rings exactly.

:::

Kernel $k$-means also has an important connection to a clustering method discussed later in this chapter: with an appropriately chosen (graph-based) kernel, minimizing the kernel $k$-means loss is equivalent to minimizing a normalized graph cut, which is exactly the objective used by spectral clustering (\@ref(sec-spec-clustering)) [@dhillon_kkmeans].  In other words, kernel $k$-means and spectral clustering are two algorithmic routes to the same underlying idea.

<!-- Hierarchical -->
```{r child = 'topics/Hierarchical-Clustering.Rmd'}
```

<!-- Model based -->
```{r child = 'topics/Model-Based-Clustering.Rmd'}
```

<!-- Spectral Clustering -->
```{r child = 'topics/Spectral-Clustering.Rmd'}
```

<!-- Clustering comparisons and metrics -->

