---
output:
  pdf_document: default
  html_document: default
---
```{r isomap-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, cache = TRUE)
set.seed(1)
require("dimRed")
require("energy")
require("scatterplot3d")
require("rgl")
require("plotly")
library("Rdimtools")
require("princurve")

myColorRamp <- function(colors, values) {
    v <- (values - min(values))/diff(range(values))
    x <- colorRamp(colors)(v)
    rgb(x[,1], x[,2], x[,3], maxColorValue = 255)
}
```


## Isometric Feature Map (ISOMAP)

### Introduction

The first manifold learning method we are going to cover is the Isometric Feature Map (ISOMAP), originally published by Tenenbaum, de Silva, and Langford in 2000 [@isomap]. As suggested by the name, we will see that the assumption of isometry is central to this method. ISOMAP combines the major algorithmic features of PCA and MDS --- computational efficiency, global optimality, and asymptotic convergence guarantees. Thanks to these extraordinary features, ISOMAP is capable of learning a broad class of nonlinear manifolds. 

### Key Definitions

**Different notions of pointwise distance**

Prior to discussing the ISOMAP algorithm, let's briefly discuss the notion of isometry through an example which motivates different notions of distance between two points.

::: {.example #ex-helix-distance name="Distance between points on a Helix"}
Consider the helix map $\Psi:\mathbb{R}\to\mathbb{R}^3$ given by the formula
\begin{equation}
\Psi(t) = \begin{bmatrix}
\frac{1}{\sqrt{2}}\cos(t) \\
\frac{1}{\sqrt{2}}\sin(t) \\
\frac{1}{\sqrt{2}}t \\
\end{bmatrix}
\end{equation}
Below, we show the result of applying the Helix map to each point in the interval $(0,25)$.  Let's focus on two points $\vec{x}_1 = \Psi(2\pi)= (1/\sqrt{2},0,\sqrt{2}\pi)^T$ and $\vec{x}_2 = \Psi(4\pi)=(1/\sqrt{2},0,2\sqrt{2}\pi)^T$ in particular which are shown as large black dots in the figure below.

```{r, echo = FALSE, message = FALSE, warning=FALSE}
library("threejs", quietly = T)
N <- 1e4
t <- sort(runif(N, min = 0, max = 25))
helix <- matrix(NA,nrow = length(t), ncol = 3)
for (n in 1:length(t)){
  helix[n,] <- c( 2^-0.5*cos(t[n]), 2^-0.5*sin(t[n]) , 2^-0.5*t[n])
}
# par(mfrow = c(1,2))
# plot(t,rep(0,N),
#      xlab = 'z', ylab = '',
#      main = "Low-Dimensional Samples",
#      cex = 0.1)
# points(c(2*pi,4*pi),rep(0,2), col = "red",cex = 1)
black_points <- data.frame(
  x = 2^-0.5*c(1,1),  
  y = 2^-0.5*c(0,0),  # Replace with your desired y coordinates
  z = 2^-0.5*pi*c(2,4)  # Replace with your desired z coordinates
)

black_points <- as.matrix(black_points)

s <- scatterplot3js(helix, color = geoelectrics::myColorRamp(c("red","purple","blue","green","yellow"), t),
              xlab = expression(x[1]), ylab = expression(x[2]), zlab = expression(x[3]),
              angle = 90,
              main = "The Helix Map",
              pch = '.',
              size = 0.1) 

threejs::points3d(s, x = black_points,  color = "black", pch = '.', size = 0.15)
```


There are a few different ways we could measure the distance between the two black points.  The first approach would be to ignore the helix (manifold) structure viewing them as vectors in $\mathbb{R}^3$ and directly measure their **Euclidean distance** which gives $$\|\vec{x}_1 - \vec{x}_2\| = \sqrt{2}\pi.$$  However, we also know that these points are images of the one-dimensional coordinate $z_1 = 2\pi$ and $z_2 = 4\pi$ respectively.  Thus, we could also consider the Euclidean distance of the lower-dimemsional coordinates which is $|2\pi - 4\pi| = 2\pi$, which notably differs from the Euclidean distance.  

A third option is to return to the three-dimensional representation but to also account for the manifold structure when considering distance. Recall Euclidean distance gives the length of the shortest, **straightline** path connecting the two points. Instead, let's restrict ourselves to only those paths which stay on the helix (manifold).  You may correctly conclude that the curve starting at $\Psi(2\pi)$, rotating up the helix one rotation, and ending at $\Psi(4\pi)$ is the shortest such path. Fortunately, computing arc-length is relatively friendly in this example since $\Psi$ already parameterizes the path connecting these two points. The arc-length is then 
$$\int_{2\pi}^{4\pi} \left\|\frac{d\Psi}{dt}\right\| dt = \int_{2\pi}^{4\pi} dt = 2\pi.$$
Jumping slightly ahead, we then say the manifold distance between $\Psi(2\pi)$ and $\Psi(4\pi)$ is $2\pi$.  Importantly, the manifold distance coincides exactly with the Euclidean distance between the **lower-dimensional** coordinates.  In fact, for any two points, $s$ and $t$, on the real line their Euclidean distance, $|s-t|$ will be the same as the manifold distance between $\Psi(s)$ and $\Psi(t)$. Thus, the helix map $\Psi$ above serves as our first example of an isometric (distance preserving) map.
:::

We may generalize this idea to any smooth manifold to define a new notion of distance.  Given a manifold $\mathcal{M}$, we define the manifold distance function $d_\mathcal{M} : \mathcal{M} \times \mathcal{M} \to [0,\infty)$ as follows 

**Definition of Manifold Distance**
::: {.definition #def-manifold-dist name="Manifold Distance Function"}
Given two points $\vec{x}$ and $\vec{y}$ on a smooth manifold, $\mathcal{M}$, let $\Gamma(\vec{x},\vec{y})$ be the set of all piecewise smooth curves connecting $\vec{x}$ and $\vec{y}$ constrained to stay on $\mathcal{M}$.  Then, we define the manifold distance to be
\begin{equation}
d_\mathcal{M}(\vec{x},\vec{y}) = \inf_{\gamma \in \Gamma(\vec{x},\vec{y})} L(\gamma)
(\#eq:def-manifold-dist)
\end{equation}
where $L(\gamma)$ is the arclength of $\gamma.$  
:::

As we reviewed above, the helix example with the arclength formula is one example of a manifold and distance function.  Additional examples of a manifold and manifold distance include, 

- Euclidean space $\mathbb{R}^d$ where standard Euclidean distance gives the manifold distance. 
- The sphere in $\mathbb{R}^3$ which is a two-dimensional manifold. Its manifold distance is also called the [Great Circle Distance](https://en.wikipedia.org/wiki/Great-circle_distance).

We may now define the notion of isometry which is a central assumption of ISOMAP.

**Definition of Isometry**
::: {.definition #def-isometry name="Isometry"}
Let $\mathcal{M}_1$ be a manifold with distance function $d_{\mathcal{M}_1}$ and let $\mathcal{M}_2$ be a second manifold with distance function $d_{\mathcal{M}_2}$.  The mapping $\Psi:\mathcal{M}_1 \mapsto \mathcal{M}_2$ is an isometry if $$d_{\mathcal{M}_1}(x,y) = d_{\mathcal{M}_2}\left(\Psi(\vec{x}),\Psi(\vec{y})\right) \qquad \text{ for all } \vec{x},\vec{y}\in \mathcal{M}_1.$$
:::

For the purposes of ISOMAP, we will think of $\mathcal{M}_1$ as some subset of a $\mathbb{R}^k$ for $k$ small where we measure distances using the Euclidean norm.  Then $\mathcal{M}_2$ will be a $k$-dimensional manifold in $\mathbb{R}^d$ containing our data. Our first assumption is that the manifold mapping $\Psi$ is an isometry. Unfortunately, in practice we do not know the manifold nor will we have a method for parameterizing curves on the manifold to compute distances. 

### Algorithm
Instead, ISOMAP makes use of a data-driven approach to estimate the manifold distance between points following a three-step procedure.  

**1) Construct Weighted Neighborhood Graph:** 

MDS uses Euclidean distance to measure pairwise distance between points $\vec{x}_i$ and $\vec{x}_j$ (data points in space $\mathcal{M}_2$), while ISOMAP uses the geodesic distance in order to reveal the underlying manifold structure. However, when the data points in the high dimensional space $\mathcal{M}_2$ have a manifold structure, usually the Euclidean pairwise distance is quite different from their pairwise geodesic distance. Fortunately, for small distances on a smoothly embedded manifold, the geodesic path between two close-by points lies nearly flat in the ambient space. So, the length of this path will be very close to the straight line (Euclidean) distance between those points in the ambient space.

The key intuition is that as the density of data points on the manifold increases (i.e., points get closer and closer), the straight line segment in the ambient space connecting two neighboring points becomes a better and better approximation of the shortest path between those points on the manifold. In the limit of the density going to infinity, these distances converge. 

Let's elucidate this concept with two illustrative examples. Firstly, imagine a two-dimensional surface, like a Swiss Roll, situated within a three-dimensional space. For an ant journeying across the Swiss Roll, the vast size difference means its immediate surroundings appear flat. From its perspective, the distance between its consecutive steps closely mirrors the distance a human might measure (Euclidean distance) – both virtually equating to the roll's geodesic distance. For a larger-scale analogy, think of Earth. Suppose extraterrestrial beings possessed technology allowing them to traverse straight through Earth's crust and mantle, thus following the shortest Euclidean path. Their journey from Los Angeles to New York might save them hundreds of miles compared to humans. However, when moving between closer landmarks, such as the Science Center to the Smith Center, their advantage diminishes.

As a result, when it comes to the measurement of geodesic distance, it is reasonable to only look at those data points that are close to each other. First, calculate all the pairwise Euclidean distance $d_{ij}=||\vec{x}_i - \vec{x}_j||_2$, then determine which points are neighbors on the manifold by connecting each point to Either (i) All points that lie within a ball of radius $\epsilon$ of that point; OR (ii) all points which are K-nearest neighbors with it. (Two different criteria, $K$ and $\epsilon$ are tuning parameters)

According to this rule, a weighted neighborhood graph $G = G(V,E)$ can be built. The set of vertices (data points in space $\mathcal{M}_2$): $V = \{\vec{x}_1, \dots , \vec{x}_N\}$ are the input data points, and the set of edges $E = \{e_{ij}\}$ indicate neighborhood relationships between the points. $e_{ij} = d_{ij}$ if (i) $||\vec{x}_i - \vec{x}_j||_2 \leq \epsilon$; OR (ii) $\vec{x}_j$ is one of the K-nearest neighbors of $\vec{x}_i$, otherwise $e_{ij} = \infty$. Sometimes, the tuning of $\epsilon$ (or $K$) is quite decisive in the output of ISOMAP, we will explain this later with a simulation example.


<!-- Fix tuning parameters $k >0$ or $\epsilon >0$.  Calculate all pairwise Euclidean distance $d_{ij} = \|\vec{x}_i - \vec{x}_j \|.$ Define a (directed) neighborhood graph $G$ by connecting nodes $i$ and $j$  -->

<!--     i) {\bf Using $k$}, connect $\vec{x}_i$ to $\vec{x}_j$ if $\vec{x}_j$ is one of the $k$ nearest points to $\vec{x}_i$ as measured by $d_{ij}$.  -->

<!--     ii) {\bf Using $\epsilon$}, connect $\vec{x}_i$ and \vec{x}_j$ if $d_ij < \epsilon.$ -->
    
    
**2) Compute graph distances**

In this step, we want to estimate the unknown true geodesic distances $\{d^{\mathcal{M}}_{ij}\}$ between all pairs of points with the help of the neighborhood graph $G$ we have just built. We use the graph distances $\{d^{\mathcal{G}}_{ij}\}$--- the shortest distances between all pairs of points in the graph $G$ to estimate $\{d^{\mathcal{M}}_{ij}\}$. For $\vec{x}_i$ and $\vec{x}_j$ that are not connected to each other, we try to find the shortest path that goes along the connected points on the graph. Following this particular sequence of neighbor-to-neighbor links, the sum of all the link weights along the path is defined as $\{d^{\mathcal{G}}_{ij}\}$. In other words, we use a number of short Euclidean distances (representing the local structure of the manifold) to approximate the geodesic distance $\{d^{\mathcal{M}}_{ij}\}$. 

This path finding step is usually done by Floyd-Warshall algorithm, which iteratively tries all transit points $k$ and find those that $\tilde{d}_{ik} + \tilde{d}_{kj} < \tilde{d}_{ij}$, and updates $\tilde{d}_{ij} = \tilde{d}_{ik} + \tilde{d}_{kj}$ for all possible combination of $i,j$. The algorithm works best in dense neighboring graph scenario, with a computational complexity of $O(n^3)$. 

The theoretical guarantee of this graph distance computation method is given by Bernstein et, al.[@Bernstein] one year after they first proposed ISOMAP in their previous paper. They show that asymptotically (as $n \rightarrow \infty$), the estimate $d^{\mathcal{G}}$ converges to $d^{\mathcal{M}}$ as long as the data points are sampled from a probability distribution that is supported by the entire manifold, and the manifold itself is flat.

The distance matrix $\Delta$ can be expressed as:
$$\Delta_{ij} = d^{\mathcal{G}}_{ij}$$

**Simulation Example**

Here we provide a randomly generated Neighborhood Graph for six data points, it uses the K-nearest neighbor criteria (can easily tell this since the matrix is not symmetric, $K=2$)

```{r}
# Define the matrix
matrix <- matrix(c(
  0,   3,  4,   Inf, Inf, Inf,
  7,   0,  Inf, 2,   Inf, Inf,
  6,   Inf,0,   Inf, 7,   Inf,
  Inf, 5,  Inf, 0,   Inf, 10,
  Inf, Inf,8,   Inf, 0,   13,
  Inf, Inf,Inf, 9,   14,  0
), byrow = TRUE, nrow = 6)
print(matrix)
```

Shown below is the implementation of Floyd-Warshall algorithm in R. As you can see from the three for loops, its computation complexity is $O(n^3)$.
```{r}
# Adjusting the matrix to set d_ij and d_ji to the smaller value
n <- dim(matrix)[1]

for (i in 1:n) {
  for (j in 1:n) {
    if (i != j && is.finite(matrix[i, j]) && is.finite(matrix[j, i])) {
      min_val <- min(matrix[i, j], matrix[j, i])
      matrix[i, j] <- min_val
      matrix[j, i] <- min_val
    }
  }
}

# Floyd-Warshall Algorithm
floyd_warshall <- function(mat) {
  n <- dim(mat)[1]
  dist <- mat
  
  for (k in 1:n) {
    for (i in 1:n) {
      for (j in 1:n) {
        dist[i, j] <- min(dist[i, j], dist[i, k] + dist[k, j])
      }
    }
  }
  
  return(dist)
}

# Get the result
result <- floyd_warshall(matrix)

# Print the result
print(result)
```

**3) Applying MDS to $\Delta$**

As mentioned before, ISOMAP can be viewed as the application of classical MDS in non-linear case. As a result, the reconstruction of $\{\vec{z}_i\}$ in the $k$ dimensional  $\mathcal{M}_1$ follows from an application of classical scaling, but we use the manifold distances estimated in the previous step. The main goal is thus to find low-dimensional coordinates which preserve the geodesic distance of the manifold in $\mathcal{M}_2$ as much as possible. 

<!-- Without any additional information, there are infinite $\{\vec{z}_i\}_{i=1}^N \subset \mathbb{R}^k$ that can be viewed as the optimal solution. For some invertible function $\Phi:\mathbb{R}^k\to\mathbb{R}^k$, a new manifold mapping $\Psi \circ \Phi^{-1}$ can be constructed. $\vec{x}_i = \Psi \circ \Phi^{-1} (\Phi(\vec{z}_i))$, which proofs that $\{\Phi(\vec{z}_i)\}$ is equivalent to $\{\vec{z}_i\}$ when it comes to the reconstruction of the lower dimensional configuration. -->

<!-- Without loss of generality, we assume that $\{\vec{z}_i\}$ are actually centered. So the distance matrix of $\{\vec{z}_i\}$ can be expressed as ${\bf B}={\bf Z Z}^T$, so that ${\bf B}_{ii}=||\vec{z}_i||^2_2$ and ${\bf B}_{ij}={\vec{z}_i}^T \vec{z}_j$. -->

<!-- The embedding vectors $\{\hat{z}_{i}\}$ (estimate of points in lower dimensional feature space $\mathcal{M}_1$) are chosen in order to minimize the objective function: -->

<!-- $$\text{argmin}_{{\bf Z}\in\mathbb{R}^{N\times k}} \|{\bf ZZ}^T - {\bf B}\|_F^2$$ -->

<!-- where the entries of ${\bf B}$ are : -->
<!-- $${\bf B}_{ij}= -\frac{1}{2} \Delta^2_{ij} + \frac{1}{d} \sum^{d}_{i=1} \Delta^2_{ij} + \frac{1}{d} \sum^{d}_{j=1} \Delta^2_{ij} - \frac{1}{2d^2} \sum^{d}_{i=1} \sum^{d}_{j=1} \Delta^2_{ij}.$$ -->

<!-- In matrix form, it is actually, $B = - \frac{1}{2} H \Delta H$, where $H = I_n - \frac{1}{n} \mathbb{1} \mathbb{1}^T$. Solving this optimization is equivalent to finding the classical scaling solution in $k$ dimensions.   -->

<!-- The next step is just a PCA problem. Implement eigen-decomposition on matrix ${\bf B}, $${\bf B}={\bf U \Lambda U}^T=  ({\bf U \Lambda}^{1/2} )({\bf U \Lambda}^{1/2} )^T$$, then arrange the singular value in descending order, find the first $k^{\prime}$ ones. We acquire $\Lambda_{k^{\prime}}$ and $U_{k^{\prime}}$.  -->
<!-- $$(\hat{z}_1 | \hat{z}_2 | \dots | \hat{z}_N) = \Lambda_{k^{\prime}} U_{k^{\prime}}$$ -->

Since we don't know the dimension of the underlying feature space, here $k^{\prime}$ is a tuning parameter. Usually, we use a scree plot ($k^{\prime}$ against the sum of the omitted eigenvalues ) and find the elbow point similar to the ideas discussed in the previous section on classical scaling.


### Limitations of ISOMAP

Though ISOMAP is a powerful manifold learning method that works well under most circumstances. It still has some limitations in certain scenarios.

1) If the data is corrupted by nonneglible noises $\{\epsilon_i\}$, then ISOMAP may fail to identify the manifold. Also, ISOMAP is quite sensitive to the tuning parameters. To alleviate the negative impact, it's highly suggested to start with a relatively small $\epsilon$ or $K$, and increase them gradually. Here we use the Swiss Roll Example to help explain this point. 

```{r}
# generate Swiss roll-shaped data
S <- rep(0,2000)
Swiss <- matrix(NA, nrow = 2000, ncol = 3)
for( n in 1:2000){
    s <- runif(1, min = 3*pi/2, max = 9*pi/2)
    t <- runif(1, min = 0,      max = 15)
    S[n] <- s
    Swiss[n, ] <- c( s*cos(s), t, s*sin(s) )
}

par(mfrow = c(2,2))
# K = 20
scatterplot3d(Swiss, color = myColorRamp(c("red","purple","blue","green","yellow"), S ),
               xlab = expression(x[1]), ylab = expression(x[2]), zlab = expression(x[3]))
plot(embed(Swiss, "Isomap", .mute = c("message", "output"), ndim =2, knn = 20)@data@data,
     xlab = "1st Dimension", ylab = "2nd Dimension", main = "K = 20",
     col = myColorRamp(c("red","purple","blue","green","yellow"), S)  
     )
# K = 50
scatterplot3d(Swiss, color = myColorRamp(c("red","purple","blue","green","yellow"), S ),
               xlab = expression(x[1]), ylab = expression(x[2]), zlab = expression(x[3]))
plot(embed(Swiss, "Isomap", .mute = c("message", "output"), ndim =2, knn = 50)@data@data,
     xlab = "1st Dimension", ylab = "2nd Dimension", main = "K = 50",
     col = myColorRamp(c("red","purple","blue","green","yellow"), S)  
     )
```

Obviously, ISOMAP performs well when $K$ is small. However, as $K$ increases, the algorithm no longer recovers the lower dimensional feature space. Because the distance between nearby arms is not large, some points from different arms are considered as "close". The pairwise geodesic distances between these points is approximated by their Euclidean distance, which is not correct, of course. 

2) When data points are sparse in certain areas or directions of the manifold, the integrity of the learned manifold structure can be compromised. The following example will clarify this notion:

Consider the Swiss Roll as an illustration. This time, however, the Swiss Roll is made sparse along its horizontal direction. To the naked eye, this sparse region seems to have minimal impact on the overall manifold. Naturally, we still perceive it as a Swiss Roll.

```{r}
set.seed(42)
S <- rep(0,2000)
Swiss_sparse <- matrix(NA, nrow = 2000, ncol = 3)
for( n in 1:2000){
    s <- runif(1, min = 3*pi/2, max = 9*pi/2)
    t <- runif(1, min = 0,      max = 60)
    S[n] <- s
    Swiss_sparse[n, ] <- c( s*cos(s), t, s*sin(s) )
}
# 
# # Manually create a sparse region
# mask <- Swiss[,1] > 2 & Swiss[,1] < 8 & Swiss[,2] > 10 & Swiss[,2] < 15
# Swiss_sparse <- Swiss[!mask, ]
# S <- S[!mask]

scatterplot3d(Swiss_sparse, color = myColorRamp(c("red","purple","blue","green","yellow"), S), main = "Swiss Roll with Sparse Region",
               xlab = expression(x[1]), ylab = expression(x[2]), zlab = expression(x[3]))
```

However, when using ISOMAP to recover the lower-dimensional feature space, complications arise, even when adjusting various tuning parameters.

```{r}
set.seed(77)
par(mfrow = c(2,2))
# K = 10
plot(embed(Swiss_sparse, "Isomap", .mute = c("message", "output"), ndim =2, knn = 10)@data@data,
     xlab = "1st Dimension", ylab = "2nd Dimension", main = "K = 10",
    col = myColorRamp(c("red","purple","blue","green","yellow"), S))
# k = 20
plot(embed(Swiss_sparse, "Isomap", .mute = c("message", "output"), ndim =2, knn = 20)@data@data,
     xlab = "1st Dimension", ylab = "2nd Dimension", main = "K = 20",
    col = myColorRamp(c("red","purple","blue","green","yellow"), S))
# K = 50
plot(embed(Swiss_sparse, "Isomap", .mute = c("message", "output"), ndim =2, knn = 50)@data@data,
     xlab = "1st Dimension", ylab = "2nd Dimension", main = "K = 50",
    col = myColorRamp(c("red","purple","blue","green","yellow"), S))
# K = 100
plot(embed(Swiss_sparse, "Isomap", .mute = c("message", "output"), ndim =2, knn = 100)@data@data,
     xlab = "1st Dimension", ylab = "2nd Dimension", main = "K = 100",
    col = myColorRamp(c("red","purple","blue","green","yellow"), S))
```


3) One of the two major assumptions of ISOMAP is that $\mathcal{M}_1$ is convex, that is to say, if the low-dimensional manifold contains many holes and concave margins (so that the manifold distance in $\mathcal{M}_1$ is not the same as the Euclidean distance) then ISOMAP will probably be not ideal. 

Here we use the example of a folded washer manifold to illustrate this point. The hole in the center of the washer breaks convexity. No matter how we tune the parameter $K$, ISOMAP always fails to recover the lower dimensional feature.

```{r}
# generate washer in 2d 
N <- 1e5
washer <- matrix(NA, nrow =N, ncol = 2)
for (n in 1:N){
  r <- runif(1,min=5,max = 10); 
  a = runif(1,min = 0, max = 2*pi); 
  washer[n,] <- r*c(cos(a),sin(a)) + c(20,0)
  }

# generate folded washer in 3D
washer3 <- cbind(washer, washer[,2]^2)

# generate rolled washer in 3D
washer.swiss <- washer3
for (n in 1:dim(washer3)[1] ){
  washer.swiss[n,] <- c( washer[n,1]*cos(washer[n,1]), washer[n,2], washer[n,1]*sin(washer[n,1] )) 
}
```

```{r}
N <- 2000
par(mfrow = c(2,2))
# K = 10
scatterplot3d(washer3[1:N,], color = myColorRamp(c("red","purple","blue","green","yellow"), washer[1:N,1] ),
               xlab = expression(x[1]), ylab = expression(x[2]), zlab = expression(x[3]))
plot(embed(washer3[1:N,], "Isomap", .mute = c("message", "output"), ndim =2, knn = 10)@data@data %*% matrix(c(0,1,1,0),nrow = 2),
     xlab = "1st Dimension", ylab = "2nd Dimension", main = "K=10",
     col = myColorRamp(c("red","purple","blue","green","yellow"), washer[,1])  )

# K = 20
plot(embed(washer3[1:N,], "Isomap", .mute = c("message", "output"), ndim =2, knn = 20)@data@data %*% matrix(c(0,1,1,0),nrow = 2),
     xlab = "1st Dimension", ylab = "2nd Dimension", main = "K=20",
     col = myColorRamp(c("red","purple","blue","green","yellow"), washer[,1])  )

# K = 50
plot(embed(washer3[1:N,], "Isomap", .mute = c("message", "output"), ndim =2, knn = 50)@data@data %*% matrix(c(0,1,1,0),nrow = 2),
     xlab = "1st Dimension", ylab = "2nd Dimension", main = "K=50",
     col = myColorRamp(c("red","purple","blue","green","yellow"), washer[,1])  )
```




<!-- 3) \underline{Embedding via MDS:}  The distances found in step 2 provide a distance matrix in which each entry approximates the manifold distance.  Under the assumption that the unknown manifold map is an isometry, we can seek a lower-dimensioal collection of points which have these pairwise distance.  By default, we use classical scaling to find such a configuration.  As discussed in Chapter \@ref(sec-mds), we have a f -->

<!--     - For visualization, we may typically choose a dimension (usually one, two, or three) and find the corresponding classical MDS solution. -->

<!--     - Alternatively, we use the eigenvalues of the inner product matrix based on the graph distances to choose a suitable cutoff for the appropriate  -->

    



