
```{r tsne-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, cache = TRUE)
set.seed(1)
require("Rtsne")
require("scatterplot3d")

myColorRamp <- function(colors, values) {
    v <- (values - min(values))/diff(range(values))
    x <- colorRamp(colors)(v)
    rgb(x[,1], x[,2], x[,3], maxColorValue = 255)
}
```

## t-Distributed Stochastic Neighbor Embedding (t-SNE) {#sec-tsne}

### Introduction

t-Distributed Stochastic Node Embedding (t-SNE) was introduced by van der Maaten and Hinton in 2008 [@tsne] and is one of the two most widely used tools for visualizing high-dimensional data.  We'll cover the other most common method, UMAP, in the next section. Both t-SNE and UMAP share a common setup.  First, build an affinity between points in $\mathbb{R}^d$, build a comparable affinity between candidate points in $\mathbb{R}^t$, and move the candidate points to make the two affinities agree.  However, they offer a more dynamic method of locally adapting the affinity for each observation. The interesting differences between the methods are in the details of how each affinity is built and what notion of "agreement" each one optimizes.

### Algorithm

#### Affinities in the original space

Like Laplacian eigenmaps, we begin by turning Euclidean distances in $\mathbb{R}^d$ into affinities. For each $\vec{x}_i$, t-SNE defines a conditional probability 
$$p_{j|i} = \frac{\exp\left(-\|\vec{x}_i-\vec{x}_j\|^2/2\sigma_i^2\right)}{\sum_{k\ne i} \exp\left(-\|\vec{x}_i-\vec{x}_k\|^2/2\sigma_i^2\right)}, \qquad p_{i|i}=0,$$
using a Gaussian centered at $\vec{x}_i$ with its own bandwidth $\sigma_i$. If we organize these probabilities into a matrix ${\bf P}$ such that ${\bf P}_{ij}=p_{j|i}$, then ${\bf P}={\bf D}^{-1}{\bf A}$ where ${\bf D}$ and ${\bf A}$ are the degree and affinity matrices from Laplacian eigenmaps except for one critical difference.  Laplacian eigenmaps use a global bandwidth $\sigma$ in the rbf kernel to compute the affinities. For t-SNE, we are using a data-dependent rbf kernel function where the bandwidth changes row-by-row.

To avoid the challenges of tuning each $\sigma_i$, van der Maaten and Hinton proposed a scheme of automatically choosing these parameters based on a globally defined tuning parameter they called *perplexity.* Let $$I(\{p_{j|i}\}_{j\ne i}) = -\sum_{j\ne i}p_{j|i}\log(p_{j|i})$$ denote the Shannon entropy of the conditional probability $\{p_{j|i}\}_{j\ne i}$.  The user specifies a perplexity, hereafter denoted as *Perp*, such that 
\begin{equation}
2^{I(\{p_{j|i}\}_{j\ne i})} = Perp.
(\#eq:target-perplexity)
\end{equation}
Each $\sigma_i$ is optimized (typically through a process called binary search) until \@ref(eq:target-perplexity) is satisfied for each $\{p_{j|i}\}_{j\ne i}$.

As a result, points in dense regions get a small $\sigma_i$ and points in sparse regions get a large one, so that every point ends up with roughly the same amount of "neighborhood mass" regardless of local density. Heuristically, you can view perplexity as a stand-in for the number of effective neighbors of $\vec{x}_i$. Increasing (decreasing) perplexity increases (decreases) neighborhood size and subquently increases (decreases) the local bandwidths.  

Next, we symmetrize and normalize the conditional probabilies,
$$p_{ij} = \frac{p_{j|i}+p_{i|j}}{2N}, \qquad 1\le i,j\le N$$
gives a discrete probability distribution.

#### Affinities in the embedding and the crowding problem

Given candidate low-dimensional points $\vec{y}_1,\dots,\vec{y}_N\in\mathbb{R}^t$, the most direct approach would define an analogous Gaussian affinity $q_{ij}$ in the embedding and ask that $q_{ij}$ match $p_{ij}$ as closely as possible. This is essentially what earlier methods (including the original stochastic node embedding that t-SNE improves on) did, but it runs into a real geometric obstruction.

Think about how much "room" there is around a point at a given distance $r$ in $\mathbb{R}^d$: the volume of a thin shell of radius $r$ grows like $r^{d-1}$, so in high dimensions there is an enormous amount of space available at moderate distances -- room enough for many points to sit at a similar, moderate similarity to $\vec{x}_i$. A two- or three-dimensional embedding has nowhere near this much room. If we insist on reproducing every moderate high-dimensional affinity with an equally moderate low-dimensional distance, the embedding is forced to crowd unrelated points close together simply because there isn't enough space to spread them all out -- distinct clusters end up squeezed into the middle of the picture. This is known as the **crowding problem**.

t-SNE's fix is to give the embedding a different, heavier-tailed notion of affinity than the original space uses. Instead of a Gaussian, we use a Student-t distribution with one degree of freedom (a Cauchy distribution),
$$
q_{ij} = \begin{cases}
0 & i=j \\
\frac{\left(1+\|\vec{y}_i-\vec{y}_j\|^2\right)^{-1}}{\sum_{\ell=1}^N\sum_{k=1}^N \left(1+\|\vec{y}_k-\vec{y}_\ell\|^2\right)^{-1}} & i\ne j 
\end{cases}
$$
The Cauchy density decays polynomially ($\propto \|\vec{y}_i-\vec{y}_j\|^{-2}$) rather than the Gaussian's exponential decay, so a moderate affinity $p_{ij}$ can now be matched by $q_{ij}$ at a *larger* embedded distance than a same-shaped Gaussian would have allowed. Geometrically, this allows moderately dissimilar points to drift apart in the embedding while genuinely close neighbors with large affinities -- which still require a small $\|\vec{y}_i-\vec{y}_j\|$ to produce a large $q_{ij}$ -- remain tightly packed. The mismatch between the two tail shapes is doing real work here, actively pushing weakly related points apart to make room for the points that matter.

#### Optimizing the embedding

With both sets of affinities in hand, t-SNE chooses $\vec{y}_1,\dots,\vec{y}_N$ to minimize the Kullback-Leibler divergence of $p_{ij}$ from $q_{ij}$, namely, we seek a configuration $\vec{y}_{1},\dots, \vec{y}_N$ minimizing
\begin{equation}
C(\vec{y}_1,\dots,\vec{y}_N) = \sum_{i=1}^N\sum_{j\ne i} p_{ij}\log\left(\frac{p_{ij}}{q_{ij}}\right).
(\#eq:t-SNE-KL)
\end{equation}
Unlike the preceding methods in this chapter, \@ref(eq:t-SNE-KL) does not have a clean analytic optima.  Instead, advanced gradient based optimization is used. Rather than delve into the details of the optimization scheme, let's take a moment investigate what gradients tell about \@ref(eq:t-SNE-KL).  Note, the gradient of $C$ w.r.t. $\vec{y}_i$ is,

$$\nabla_{\vec{y}_i} C(\vec{y}_1,\dots,\vec{y}_N) = 4\sum_{j\ne i} \frac{p_{ij}-q_{ij}}{1+\|\vec{y}_i - \vec{y}_j\|^2}(\vec{y}_i - \vec{y}_j).$$
This gradient is asymmetric in a useful way: it penalizes using a small $q_{ij}$ to represent a large $p_{ij}$ far more heavily than the reverse, so the optimization prioritizes keeping genuine neighbors close over faithfully preserving large distances. 
Pairs with $p_{ij} > q_{ij}$ pull $\vec{y}_i$ and $\vec{y}_j$ together, while pairs with $p_{ij} < q_{ij}$ push them apart. 

In practice, the optimization is initialized (often from a small random configuration or a PCA projection), run with momentum to speed convergence, and typically uses an "early exaggeration" phase which temporarily inflates the $p_{ij}$ to encourage well-separated clusters to form before fine details are refined. Interested readers should visit the original paper for additional details.

::: {.example #tsne-swiss name="t-SNE applied to the Swiss Roll"}

Let's revisit the swiss roll one more time and compare t-SNE's embedding to what we've seen from ISOMAP, LLE, and UMAP.

```{r, echo = FALSE, fig.cap = "The Swiss Roll (left) and its t-SNE embedding (right)"}
N <- 2000
s <- runif(N, min = pi/2, max = 9*pi/2)
t <- runif(N, min = 0, max = 15)
swiss <- cbind(s*sin(s), s*cos(s), t)
embedding <- Rtsne(swiss, dims = 2, perplexity = 30, check_duplicates = FALSE)$Y
par(mfrow = c(1,2))
scatterplot3d(swiss, color = myColorRamp(c("red","purple","blue","green","yellow"), s),
              xlab = "", ylab = "", zlab = "", main = "Swiss Roll", cex.symbols = 0.3, angle = 90)
plot(embedding[,1], embedding[,2], col = myColorRamp(c("red","purple","blue","green","yellow"), s),
     xlab = "", ylab = "", main = "t-SNE Embedding", pch = 16, cex = 0.5)
```

t-SNE successfully separates the roll along its length, and the coloring confirms that nearby points on the manifold remain nearby in the embedding. The result is not a faithful rectangle -- t-SNE has stretched, compressed, and torn the manifold into disconnected pieces in a few places, a direct consequence of the crowding problem discussion above: the algorithm is actively trading distance-fidelity for the breathing room needed to keep local neighborhoods intact.

:::

### Strengths and Weaknesses

#### Strengths

1) **Excellent at revealing cluster structure**

The combination of the asymmetric KL loss and the heavy-tailed embedding affinity makes t-SNE particularly good at pulling apart genuinely distinct clusters, which is a large part of why it became a default first step for visualizing datasets.

2) **Few conceptual moving parts**

Beyond the perplexity (conventionally chosen between five and fifty), there are relatively few choices to make, and the perplexity itself has a fairly direct interpretation (an effective neighborhood size) that makes it easier to reason about than some of the tuning parameters of other methods.

3) **Directly addresses a real geometric obstruction**

Unlike methods that inherit a single fixed notion of similarity from the ambient space, t-SNE's use of mismatched tail behavior between $p_{ij}$ and $q_{ij}$ is a deliberate, principled response to the crowding problem, not an incidental design choice. Additionally, the locally adaptive choice of bandwidths is better suited to large variations in sampling density along the manifold containing the data.

#### Weaknesses

1) **Distances and cluster sizes are not meaningful**

t-SNE prioritizes neighborhoods over distances. The size of a cluster in a t-SNE plot and the gap between two clusters are both driven largely by the optimization and the perplexity, not by the true geometry of the data, and should not be over-interpreted.

2) **No natural out-of-sample extension**

t-SNE optimizes the positions of a fixed set of points directly; there is no manifold map $\Psi$ or its inverse to apply to a new observation, so embedding new data typically requires rerunning the entire optimization (or falling back on a separate out-of-sample heuristic).

3) **Sensitive to perplexity, and computationally demanding**

Very different perplexities can produce visually different embeddings of the same data. t-SNE's optimization has no principled starting point, so different runs (or different random initializations) can converge to different local optima since \@ref(eq:t-SNE-KL) is not convex.  The $O(N^2)$ cost of the exact pairwise affinities above also makes t-SNE slow for large datasets in practice, though accelerated approximations are commonly used to mitigate this issue. 

